---
title: "gini coefficient"
output: html_document
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---
```{r}
library(dplyr)
library(DescTools)
library(lctools)
library(tidycensus)
```

# total population
```{r}

options(tigris_use_cache = TRUE)
years <- c(2009:2023) # 2009 
# Initialize an empty dataframe to store results

total_pop_data <- data.frame()
# Loop through each year and state
for (year in years) {
  for (state_code in  state.abb) {
    message(paste("Fetching data for", state_code, "in", year))
    
    # Try to fetch the data (wrap in tryCatch in case a year/state has issues)
    tryCatch({
      totalpop <- get_acs(
        geography = "tract", 
        variables = "B01001_001E", 
        state = state_code,
        year = year, 
        survey = "acs5", 
        output = "wide"
      )
      
      # Add year and state as columns
      totalpop$Year <- year
      totalpop$State <- state_code
      
      # Append to the main dataframe
      total_pop_data <- bind_rows(total_pop_data, totalpop)
    }, error = function(e) {
      warning(paste("Failed for", state_code, "in", year, ":", e$message))
    })
  }
}


# Check the resulting dataframe
total_pop_data<- total_pop_data%>%select(-c("B01001_001M"))%>%  rename(total_pop=B01001_001E,year = Year)
write.csv(total_pop_data,'GINI Paper/Data/Tractlevel_totalpop_acs_2009_2023.csv', row.names=FALSE)
```


# employment  (need to redo it at census tract level)
```{r}
# population data block group level
total_pop_data <- read.csv('Tractlevel_totalpop_acs_2009_2023.csv')
total_pop_data$GEOID <- sprintf("%011s",total_pop_data$GEOID)

# employment data block level
df <- read.csv('GiniData.csv')%>%  rename(Others=Accommodation...Food.Services,BlockGEOID= w_geocode, State=state)%>%select(-c(X))

# capitalizing state abb 
df$State<- toupper(df$State) 

# chaing block group code to have a leading zero
df$BlockGEOID<- sprintf("%015.0f", as.numeric(df$BlockGEOID))

# extracting 11 digits which indicates census tract level
df$GEOID <- substr(df$BlockGEOID, 1, 11)

# group by to census block group level
DF <- df %>%
  group_by(GEOID, year) %>%
  summarise(
    Educational_Services = sum(Educational.Services, na.rm = TRUE),
    Health_Care = sum(Health.Care, na.rm = TRUE),
    Others = sum(Others, na.rm = TRUE),
    .groups = 'drop'
  )%>%filter(year>=2009)%>%full_join(total_pop_data)%>%select(-c("NAME")) # Filter rows based on condition


write.csv(DF,'GINI Paper/Data/Tractlevel_Employment_By_Industry_withTotalpop_2009_2023.csv', row.names=FALSE)
```

# calculating gini
```{r}
# Extract first 11 digits (Census tract)
DFfilter$tract_GEOID <- substr(DFfilter$GEOID, 1, 11)

# Count unique tracts
n_unique_tracts <- length(unique(DFfilter$tract_GEOID))
print(n_unique_tracts)
```

# Code with Parallel Processing
```{r}
library(foreach)
library(doParallel)

DF <- read.csv('Tractlevel_Employment_By_Industry_withTotalpop_2009_2023.csv')
DF$GEOID <- sprintf("%011s", DF$GEOID)
DF$GEOIDCN<-  substr(DF$GEOID, 1, 5)

# Filter rows based on condition
# Define the types of interest
types <- c("Health_Care", "Educational_Services", "Others")
counties <- unique(DF$GEOIDCN)

# Ensure 'year' is numeric if not already converted
DF$year <- as.numeric(DF$year)
unique_years <- unique(DF$year)

# make it as a long table
DF_long <- as.data.frame(DF) %>%
  pivot_longer(
    cols = c(Educational_Services, Health_Care, Others),
    names_to = "TYPE",
    values_to = "IndustryEmployment"
  )%>%
  mutate(total_pop = ifelse(is.na(total_pop), 0, total_pop))%>%
  mutate(IndustryEmployment = ifelse(is.na(IndustryEmployment), 0, IndustryEmployment))


# Set up parallel backend
num_cores <- parallel::detectCores() - 4  # Leave 1 core free
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Create a dataframe with all combinations
combo_df <- expand.grid(
  Year = unique_years,
  Type = types,
  County = counties
) # Filter rows based on condition


# Initialize an empty dataframe to store results
gini_results <- data.frame(County = character(), Year = integer(), Type = character(), Gini = numeric(),lwr.ci =numeric(),   upr.ci=numeric(),  stringsAsFactors = FALSE)


# Run parallel processing
gini_results <- foreach(i = 1:nrow(combo_df), .combine = rbind, .packages = c("dplyr", "DescTools")) %dopar% {
  current_year <- combo_df$Year[i]
  type <- combo_df$Type[i]
  county <- combo_df$County[i]

  # Filter the data
  DFfilter <- DF_long %>%
    filter(year == current_year, TYPE == type, GEOIDCN == county) %>%
    filter(!is.na(IndustryEmployment), !is.na(total_pop), total_pop != 0)

# Safeguard for empty or invalid data
  if (!is.null(DFfilter) && nrow(DFfilter) > 1 && all(c("IndustryEmployment", "total_pop") %in% names(DFfilter))) {
    gini_value <- tryCatch(
      {
        Gini(weights = DFfilter$total_pop, DFfilter$IndustryEmployment,  unbiased = FALSE) # Filter rows based on condition
      },
      error = function(e) {
        c(NA, NA, NA)
      }
    )
  } else {
    gini_value <- c(NA, NA, NA)
  }

  # Return the result row
  data.frame(
    GEOIDCN = county,
    Year = current_year,
    Type = type,
    Gini = gini_value[1],
    lwr.ci = gini_value[2],
    upr.ci = gini_value[3],
    stringsAsFactors = FALSE
  )
}

# Stop parallel cluster
stopCluster(cl)

# View and reshape
print(gini_results)

gini_results_wide <- gini_results %>%
  tidyr::pivot_wider(
    id_cols = c(GEOIDCN, Year), 
    names_from = Type, 
    values_from = Gini
  ) %>%
  filter(Year != 2024)

print(gini_results_wide)

write.csv(gini_results_wide,'GINI Paper/Data/County_gini_coefficient.csv', row.names=FALSE)
```

# visualization
```{r}
library(ggplot2)
library(geofacet)
library(fmsb)
library(tidycensus)
```

```{r}
gini_results_wide <- read.csv('County_gini_coefficient.csv')
gini_results_wide$GEOIDCN<-  sprintf("%05d", as.integer(gini_results_wide$GEOIDCN))

gini_results_wide$State<-  substr(gini_results_wide$GEOIDCN, 1, 2)

state_info<- fips_codes %>%
  select(state, state_code) %>%
  distinct() %>%
  rename(
    State_Abbreviation = state,
    State = state_code
  )



```

# geo facet map (Figure 3)
```{r}
library(tidyr)
library(ggplot2)
summarydf <- gini_results_wide %>%
  left_join(state_info, by = "State") %>%
  filter(!State %in% c("66", "72", "78")) %>%  # Filter rows based on condition
  filter(!is.na(State_Abbreviation)) %>%
  group_by(State, State_Abbreviation) %>%
  summarise(
    Health_Care = mean(Health_Care, na.rm = TRUE),
    Educational_Services = mean(Educational_Services, na.rm = TRUE),
    Others = mean(Others, na.rm = TRUE),
    .groups = "drop"
  )

summarydf_long <- summarydf %>%
  pivot_longer(cols = c("Health_Care", "Educational_Services", "Others"), names_to = "Category", values_to = "Value")%>%group_by(State, Category, State_Abbreviation )%>%summarise(Meangini = mean(Value))

ggplot(summarydf_long, aes(x = Category, y = Meangini, fill = Category)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  facet_geo(~ State_Abbreviation, grid = "us_state_grid2") +  # Adjust grid as needed
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Statewise Comparison of Health Metrics", y = "Index Value", x = "")

ggplot(summarydf_long, aes(x = Category, y = Meangini, fill = Category)) +
  geom_bar(stat = "identity", show.legend = TRUE, width = 0.7) +  # Use geom_bar
  coord_polar() +  # This converts the bar chart to a radial chart
  facet_geo(~ State_Abbreviation, grid = "us_state_grid1") +  # Adjust grid as needed
  theme_minimal() +
  theme(axis.text.x = element_blank()) 
  # labs(title = "Statewise Comparison of Health Metrics", y = "Index Value", x = "")
```

# timeserise (Figure 3)
```{r}
library(dplyr)
library(tidyr)

summarydf <- gini_results_wide %>%
  left_join(state_info, by = "State") %>%
  filter(!State %in% c("66", "72", "78")) %>%  # Filter rows based on condition
  filter(!is.na(State_Abbreviation)) %>%
  group_by(State, State_Abbreviation, Year) %>%
  summarise(
    Health_Care = mean(Health_Care, na.rm = TRUE),
    Educational_Services = mean(Educational_Services, na.rm = TRUE),
    Others = mean(Others, na.rm = TRUE),
    .groups = "drop"
  )

summarydf_long <- summarydf %>%
  pivot_longer(
    cols = c("Health_Care", "Educational_Services", "Others"),
    names_to = "Category",
    values_to = "Value"
  ) %>%
  mutate(Category = recode(Category,
    "Health_Care" = "Health Care",
    "Educational_Services" = "Educational Services",
    "Others" = "Daily Living"
  )) %>%
  mutate(Category = factor(Category, levels = c("Daily Living", "Educational Services","Health Care" )))

category_colors <- c(
    "Daily Living" = "steelblue",
    "Educational Services" = "firebrick",
    "Health Care" = "forestgreen"
)


ggplot(summarydf_long, aes(x = Year, y = Value, color = Category, group = Category)) +
  geom_line(size = 0.8) +
  facet_geo(~ State_Abbreviation, grid = "us_state_grid1", scales = "free_x") +
  scale_color_manual(values = category_colors) +
  labs(
    x = "Year",
    y = "Value",
    color = "Category"
  ) +
  scale_x_continuous(breaks = seq(2009, 2025, 5)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title = element_blank(),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12)
  )



ggsave('GINI Paper/Draft/Figures/County_gini_coefficient_timeseries.png',dpi=300, width = 13, height = 9)
```


# Location quotient
```{r}
library(dplyr)
```

```{r}
df <- read.csv('GiniData.csv')%>%  rename(Others=Accommodation...Food.Services,BlockGEOID= w_geocode, State=state)%>%select(-c(X))
# chaing block group code to have a leading zero
df$BlockGEOID<- sprintf("%015.0f", as.numeric(df$BlockGEOID))
df$GEOIDCN<-  substr(df$BlockGEOID, 1, 5)

DF <- df %>%
  group_by(GEOIDCN, year) %>%
  summarise(
    Educational_Services = sum(Educational.Services, na.rm = TRUE),
    Health_Care = sum(Health.Care, na.rm = TRUE),
    Others = sum(Others, na.rm = TRUE),
    total_employment = sum(total_employment, na.rm=TRUE),
    .groups = 'drop')
  
```

```{r}
calculate_lq <- function(df, industry_cols, region_col = "GEOIDCN", year_col = "year", total_col = "total_employment") {
  library(dplyr)
  
  # Step 1: Calculate national totals (by year)
  national_totals <- df %>%
    group_by(across(all_of(year_col))) %>%
    summarise(
      across(all_of(industry_cols), sum, na.rm = TRUE),
      nat_total = sum(.data[[total_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  # Step 2: Merge national totals into original data
  df_lq <- df %>%
    left_join(national_totals, by = year_col)
  
  # Step 3: Calculate Location Quotient for each industry
  for (industry in industry_cols) {
    industry_local <- paste0(industry, ".x")  # After join
    industry_national <- paste0(industry, ".y")
    
    # Compute LQ using renamed columns
    df_lq[[paste0("LQ_", industry)]] <- 
      (df_lq[[industry_local]] / df_lq[[total_col]]) /
      (df_lq[[industry_national]] / df_lq$nat_total)
  }
  
  return(df_lq)
}


```

```{r}
# Your column names for industries
industry_columns <- c("Educational_Services", "Health_Care", "Others")

# Assuming your data is in a dataframe called `data`
lq_data <- calculate_lq(DF, industry_columns )

LQdat <- lq_data %>%select(c(GEOIDCN, year, LQ_Educational_Services, LQ_Health_Care, LQ_Others))%>% rename(Year=year )


# write.csv(LQdat,'GINI Paper/Data/County_Location_Quotient.csv', row.names=FALSE)
```

# LQ geofacet (Figure 3)
```{r}
library(ggplot2)
library(geofacet)
library(fmsb)
library(tidycensus)
```

```{r}
LQdat <- read.csv('County_Location_Quotient.csv')

LQdat$GEOIDCN<-  sprintf("%05d", as.integer(LQdat$GEOIDCN))


state_info<- fips_codes %>%
  select(state, state_code) %>%
  distinct() %>%
  rename(
    State_Abbreviation = state,
    State = state_code
  )

LQdat$State<-  substr(LQdat$GEOIDCN, 1, 2)



summarydf <- LQdat %>%
  left_join(state_info, by = "State") %>%
  filter(!State %in% c("66", "72", "78")) %>%  # Filter rows based on condition
  filter(!is.na(State_Abbreviation)) %>%
  filter(Year>=2009)%>%
  group_by(State,Year, State_Abbreviation) %>%
  summarise(
    Health_Care = mean(LQ_Health_Care, na.rm = TRUE),
    Educational_Services = mean(LQ_Educational_Services, na.rm = TRUE),
    Others = mean(LQ_Others, na.rm = TRUE),
    .groups = "drop"
  )

summarydf_long <- summarydf %>%
  pivot_longer(
    cols = c("Health_Care", "Educational_Services", "Others"),
    names_to = "Category",
    values_to = "Value"
  ) %>%
  mutate(Category = recode(Category,
    "Health_Care" = "Health Care",
    "Educational_Services" = "Educational Services",
    "Others" = "Daily Living"
  )) %>%
    mutate(Category = factor(Category, levels = c("Daily Living", "Educational Services","Health Care" )))


category_colors <- c(
  "Health Care" = "forestgreen",
  "Educational Services" = "firebrick",
  "Daily Living" = "steelblue"
)


ggplot(summarydf_long, aes(x = Year, y = Value, color = Category, group = Category)) +
  # geom_hline(yintercept = 1, linetype = "dashed", color = "black") +
  geom_line(size= 0.8) +
  facet_geo(~ State_Abbreviation, grid = "us_state_grid1", scales = "free_y") +
  scale_color_manual(values = category_colors) +
  labs(# title = "Trends by State and Category",
       x = "Year",
       y = "Value",
       color = "Category") +
  scale_x_continuous(breaks = seq(2009, 2025, 5), limits = c(2009, 2023)) +
  theme_minimal() +
    theme(
      axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
      axis.text.y = element_text(size = 12),
      axis.title = element_blank(),
      legend.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12)
    )


ggsave('GINI Paper/Draft/Figures/County_LQ_timeseries.png',dpi=300, width = 13, height = 9)
```

# CARES temporal plot (Figure 3)
```{r}
library(dplyr)
library(tidyr)

care<- read.csv('County_everyyear_CareScore.csv')
care$GEOIDCN<-  sprintf("%05d", as.integer(care$GEOIDCN))

# Ensure GEOID is a 5-character string (just in case)
care <- care %>%
  mutate(GEOID = sprintf("%05s", GEOIDCN))

# Extract state FIPS from county GEOID
care <- care %>%
  mutate(State = substr(GEOID, 1, 2))

# Summarize to state-year level
care_state_year <- care %>%
  group_by(State, Year) %>%
  summarise(
    combined_score_health = mean(combined_score_health, na.rm = TRUE),
    combined_score_educ = mean(combined_score_educ, na.rm = TRUE),
    combined_score_others = mean(combined_score_others, na.rm = TRUE),
    .groups = "drop"
  )

# Add state abbreviations (assumes you have a lookup table)
state_info <- state_info %>% mutate(State = as.character(State))  # Modify or create new variables

care_state_year <- care_state_year %>%
  left_join(state_info, by = "State")

care_long <- care_state_year %>%
  pivot_longer(
    cols = starts_with("combined_score"),
    names_to = "Category",
    values_to = "Value"
  ) %>%
  mutate(
    Category = recode(Category,
      "combined_score_health" = "Health Care",
      "combined_score_educ" = "Educational Services",
      "combined_score_others" = "Daily Living"
    )
  )%>%  mutate(Category = factor(Category, levels = c("Daily Living", "Educational Services","Health Care" )))



ggplot(care_long, aes(x = Year, y = Value, color = Category, group = Category)) +
  geom_line(size = 0.8) +
  facet_geo(~ State_Abbreviation, grid = "us_state_grid1", scales = "free_y") +
  scale_color_manual(values = category_colors) +
  labs(
    x = "Year",
    y = "Combined Score",
    color = "Category"
  ) +
  scale_x_continuous(breaks = seq(2009, 2025, 5), limits = c(2009, 2023)) +
  theme_minimal() +
    theme(
      axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
      axis.text.y = element_text(size = 12),
      axis.title = element_blank(),
      legend.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12)
    )


ggsave('GINI Paper/Draft/Figures/County_CARES_timeseries.png',dpi=300, width = 13, height = 9)
```


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(cowplot)
library(tidycensus)
```

```{r}
LQdat<- read.csv('County_Location_Quotient.csv')
LQdat$GEOIDCN <- sprintf("%05s", LQdat$GEOIDCN)

gini_results <- read.csv('County_gini_coefficient.csv')
gini_results$GEOIDCN <- sprintf("%05s", gini_results$GEOIDCN)

# State abb
state_info<- fips_codes %>%
  select(state, state_code) %>%
  distinct() %>%
  rename(
    State_Abbreviation = state,
    GEOIDST = state_code
  )

DF<- left_join(gini_results, LQdat)

DF$GEOIDST <-substr(DF$GEOIDCN, 1,2)
DF <- left_join(DF,state_info)


# write.csv(DF,'GINI Paper/Data/County_GINI_LQ.csv',row.names = FALSE)
```

# visualize gini vs LQ (Figure 2)
```{r}
DF <- read.csv('County_GINI_LQ.csv')
DF$GEOIDCN <- sprintf("%05s",DF$GEOIDCN)

years <- c(2009:2021)
# Initialize an empty dataframe to store results
conus_states <- state.abb # [!state.abb %in% c("AK", "HI")]

total_pop_data <- data.frame()
# Loop through each year and state
for (year in years) {
  totalpop <- get_acs(
        geography = "county", 
        variables = "B01001_001E", 
        year = year, 
        survey = "acs5", 
        output = "wide"
      )
  # Add year and state as columns
      totalpop$Year <- year

            # Append to the main dataframe
      total_pop_data <- bind_rows(total_pop_data, totalpop)
}

total_pop_data<- total_pop_data%>%select(GEOID,B01001_001E, Year)%>% rename(GEOIDCN=GEOID ,  totalpop=B01001_001E)
```

```{r}
# Step 1: Compute ranges for x and y variables
range_x <- range(0.3,0.85)
range_y <- range(0,2, na.rm = TRUE)


plot_lq_st <- function(data, x_var, y_var, x_limits = NULL, y_limits = NULL) {
  data %>%
    filter(Year > 2008, Year < 2022) %>%
    group_by(GEOIDST, State_Abbreviation) %>%
    summarise(
      Educational_Services = mean(Educational_Services, na.rm = TRUE),
      Health_Care = mean(Health_Care, na.rm = TRUE),
      Others = mean(Others, na.rm = TRUE),
      LQ_Educational_Services = mean(LQ_Educational_Services, na.rm = TRUE),
      LQ_Health_Care = mean(LQ_Health_Care, na.rm = TRUE),
      LQ_Others = mean(LQ_Others, na.rm = TRUE),
      .groups = 'drop'
    ) %>%
    ggplot(aes_string(x = x_var, y = y_var, label = "State_Abbreviation")) +
    geom_point(color = '# 2b8cbe', alpha = 0.6) +
    ggrepel::geom_text_repel(size = 4, fontface = "bold", max.overlaps = Inf) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "# de2d26", linewidth = 1) +
    geom_vline(xintercept = 0.5, linetype = "dotted", color = "# 636363", linewidth = 1) +
    coord_cartesian(xlim = x_limits, ylim = y_limits) +
    labs(x = "Gini Coefficient", y = "Location Quotient") +
    theme_minimal() +
    theme(
      axis.title.x = element_text(size = 16, face = "bold"),
      axis.title.y = element_text(size = 16, face = "bold"),
      axis.text.x = element_text(size = 14, face = "bold"),
      axis.text.y = element_text(size = 14, face = "bold")
    )
}

# weighted pop
range_x <- range(0.3,0.9)
range_y <- range(0,2, na.rm = TRUE)


plot_lq_st <- function(data, x_var, y_var, x_limits = NULL, y_limits = NULL) {
  data %>%
    filter(Year > 2012, Year < 2022) %>%
    left_join(total_pop_data)%>%
    group_by(GEOIDST, State_Abbreviation) %>%
    summarise(
      total_pop = sum(totalpop, na.rm = TRUE),
      Educational_Services = mean(Educational_Services, na.rm = TRUE),
      Health_Care = mean(Health_Care, na.rm = TRUE),
      Others = mean(Others, na.rm = TRUE),
      LQ_Educational_Services = weighted.mean(LQ_Educational_Services, totalpop, na.rm = TRUE),
      LQ_Health_Care = weighted.mean(LQ_Health_Care,totalpop, na.rm = TRUE),
      LQ_Others = weighted.mean(LQ_Others, totalpop, na.rm = TRUE),
      .groups = 'drop'
    ) %>%
    ggplot(aes_string(x = x_var, y = y_var, label = "State_Abbreviation")) +
    geom_point(color = '# 2b8cbe', alpha = 0.6) +
    ggrepel::geom_text_repel(size = 4, fontface = "bold", max.overlaps = Inf) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "# de2d26", linewidth = 1) +
    geom_vline(xintercept = 0.5, linetype = "dotted", color = "# 636363", linewidth = 1) +
    coord_cartesian(xlim = x_limits, ylim = y_limits) +
    labs(x = "Gini Coefficient", y = "Location Quotient") +
    theme_minimal() +
    theme(
      axis.title.x = element_text(size = 16, face = "bold"),
      axis.title.y = element_text(size = 16, face = "bold"),
      axis.text.x = element_text(size = 14, face = "bold"),
      axis.text.y = element_text(size = 14, face = "bold")
    )    +
    annotate("text", x = 0.75, y = 1.8, label = "High GINI – High LQ\n(Concentrated Access)", size = 4, fontface = "bold") +
    annotate("text", x = 0.4, y = 1.8, label = "Low GINI – High LQ\n(Equitable Coverage)", size = 4, fontface = "bold") +
    annotate("text", x = 0.75, y = 0.3, label = "High GINI – Low LQ\n(Unequal Scarcity)", size = 4, fontface = "bold") +
    annotate("text", x = 0.4, y = 0.3, label = "Low GINI – Low LQ\n(Even Desert)", size = 4, fontface = "bold")

}


p1 <- plot_lq_st(DF,
                 x_var = "Health_Care",
                 y_var = "LQ_Health_Care",
                 x_limits = range_x,
                 y_limits = range_y)

p2 <- plot_lq_st(DF,
                 x_var = "Educational_Services",
                 y_var = "LQ_Educational_Services",
                 x_limits = range_x,
                 y_limits = range_y)

p3 <- plot_lq_st(DF,
                 x_var = "Others",
                 y_var = "LQ_Others",
                 x_limits = range_x,
                 y_limits = range_y)

```

```{r}
range_x <- range(0,1)
range_y <- range(0,8, na.rm = TRUE)


plot_lq <- function(data, year_range, x_var, y_var,x_limits = NULL, y_limits = NULL) {
  data %>%
    filter(Year > year_range[1], Year < year_range[2]) %>%
    ggplot(aes_string(x = x_var, y = y_var)) +
    geom_point(color = '# 2b8cbe', alpha = 0.6) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "# de2d26", linewidth = 1) +
    geom_vline(xintercept = 0.5, linetype = "dotted", color = "# 636363", linewidth = 1) +
    coord_cartesian(xlim = x_limits, ylim = y_limits) +
    labs(x ="Gini Coefficient", y = "Location Quotient") +
    theme_minimal()+
    theme(
      axis.title.x = element_text(size = 16, face = "bold"),
      axis.title.y = element_text(size = 16, face = "bold"),
      axis.text.x = element_text(size = 14, face = "bold"),
      axis.text.y = element_text(size = 14, face = "bold")
    )+
    annotate("text", x = 0.75, y = 7, label = "High GINI – High LQ\n(Concentrated Access)", size = 4, fontface = "bold") +
    annotate("text", x = 0.2, y = 7, label = "Low GINI – High LQ\n(Equitable Coverage)", size = 4, fontface = "bold") +
    annotate("text", x = 0.75, y = -0.7, label = "High GINI – Low LQ\n(Unequal Scarcity)", size = 4, fontface = "bold") +
    annotate("text", x = 0.2, y = -0.7, label = "Low GINI – Low LQ\n(Even Desert)", size = 4, fontface = "bold")
}



p4<- plot_lq(DF, c(2008, 2022),
        x_var = "Health_Care",
        y_var = "LQ_Health_Care",
        x_limits = range_x,
                 y_limits = c(-2,8)) # range_y

p5<- plot_lq(DF,  c(2008, 2022),
        x_var = "Educational_Services",
        y_var = "LQ_Educational_Services",
        x_limits = range_x,
                 y_limits = c(-2,8))

p6<- plot_lq(DF,  c(2008, 2022),
        x_var = "Others",
        y_var = "LQ_Others",
        x_limits = range_x,
                 y_limits = c(-2,8))


plot_grid(p1, p2, p3, p4, p5, p6,p7,
          labels = c("(a)", "(b)", "(c)", "(d)", "(e)", "(f)","(g)"),
          ncol = 3, nrow = 3)

ggsave('GINI Paper/Draft/Figures/County_gini_coefficient.png',dpi=300, height = 30, width = 45, units = 'cm')
```


# color for urban  (Figure 2)
```{r}
# Ensure RUCC_2013 is treated as a factor with the correct labels

# adding urbanicity
urban <- read.csv('ruralurbancodes2013.csv')
urban$GEOIDCN <-  sprintf("%05s",urban$FIPS)
urban$RUCC_2013<- as.character(urban$RUCC_2013)

DF<- DF%>%left_join(urban)

rucc_labels <- c(
  "1" = "Metro ≥1M",
  "2" = "Metro 250K–1M",
  "3" = "Metro <250K",
  "4" = "Urban Adjacent",
  "5" = "Urban Non-Adjacent",
  "6" = "Small Town Adjacent",
  "7" = "Small Town Non-Adjacent",
  "8" = "Rural Adjacent",
  "9" = "Rural Remote"
)
DF <- DF %>%
  mutate(RUCC_2013 = factor(RUCC_2013, levels = names(rucc_labels), labels = rucc_labels))

# Modified function with color by RUCC_2013
plot_lq <- function(data, year_range, x_var, y_var, x_limits = NULL, y_limits = NULL) {
  data %>%
    filter(Year > year_range[1], Year < year_range[2]) %>%
    ggplot(aes_string(x = x_var, y = y_var, color = "RUCC_2013")) +
    geom_point(alpha = 0.6, size = 2) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "# de2d26", linewidth = 1) +
    geom_vline(xintercept = 0.5, linetype = "dotted", color = "# 636363", linewidth = 1) +
    coord_cartesian(xlim = x_limits, ylim = y_limits) +
    labs(x = "Gini Coefficient", y = "Location Quotient", color = "Urbanicity") +
    scale_color_manual(values = RColorBrewer::brewer.pal(n = 9, name = "Set1")) +
    theme_minimal() +
    theme(
      axis.title.x = element_text(size = 16, face = "bold"),
      axis.title.y = element_text(size = 16, face = "bold"),
      axis.text.x = element_text(size = 14, face = "bold"),
      axis.text.y = element_text(size = 14, face = "bold"),
      legend.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12)
    ) +
    annotate("text", x = 0.75, y = 7, label = "High GINI – High LQ\n(Concentrated Access)", size = 4, fontface = "bold") +
    annotate("text", x = 0.2, y = 7, label = "Low GINI – High LQ\n(Equitable Coverage)", size = 4, fontface = "bold") +
    annotate("text", x = 0.75, y = -0.7, label = "High GINI – Low LQ\n(Unequal Scarcity)", size = 4, fontface = "bold") +
    annotate("text", x = 0.2, y = -0.7, label = "Low GINI – Low LQ\n(Even Desert)", size = 4, fontface = "bold")
}

# Create plots
p4 <- plot_lq(DF, c(2008, 2022), x_var = "Health_Care", y_var = "LQ_Health_Care", x_limits = range_x, y_limits = c(-2, 8))
p5 <- plot_lq(DF, c(2008, 2022), x_var = "Educational_Services", y_var = "LQ_Educational_Services", x_limits = range_x, y_limits = c(-2, 8))
p6 <- plot_lq(DF, c(2008, 2022), x_var = "Others", y_var = "LQ_Others", x_limits = range_x, y_limits = c(-2, 8))

```

 
# making urban and topology figure  (Figure 2)
```{r}
# making indicator

# gini >0.5, LQ >1 : inequal_havg
# gini >0.5, LQ <1 : inequal_lavg
# gini <0.5, LQ >1 : equal_havg
# gini <0.5, LQ <1 : equal_lavg
# adding urbanicity
urban <- read.csv('ruralurbancodes2013.csv')
urban$GEOIDCN <-  sprintf("%05s",urban$FIPS)
urban$RUCC_2013<- as.character(urban$RUCC_2013)

DF<- DF%>%left_join(urban)

rucc_labels <- c(
  "1" = "Metro ≥1M",
  "2" = "Metro 250K–1M",
  "3" = "Metro <250K",
  "4" = "Urban Adjacent",
  "5" = "Urban Non-Adjacent",
  "6" = "Small Town Adjacent",
  "7" = "Small Town Non-Adjacent",
  "8" = "Rural Adjacent",
  "9" = "Rural Remote"
)
DF <- DF %>%
  mutate(RUCC_2013 = factor(RUCC_2013, levels = names(rucc_labels), labels = rucc_labels))


DFindex <- DF%>%
  filter(Year > 2008, Year < 2022) %>%
  mutate(
    category_Educational = case_when(
      Educational_Services > 0.5 & LQ_Educational_Services > 1 ~ "inequal_havg",
      Educational_Services > 0.5 & LQ_Educational_Services < 1 ~ "inequal_lavg",
      Educational_Services <= 0.5 & LQ_Educational_Services > 1 ~ "equal_havg",
      Educational_Services <= 0.5 & LQ_Educational_Services < 1 ~ "equal_lavg",
      TRUE ~ NA_character_
    ),
    category_Health = case_when(
      Health_Care > 0.5 & LQ_Health_Care > 1 ~ "inequal_havg",
      Health_Care > 0.5 & LQ_Health_Care < 1 ~ "inequal_lavg",
      Health_Care <= 0.5 & LQ_Health_Care > 1 ~ "equal_havg",
      Health_Care <= 0.5 & LQ_Health_Care < 1 ~ "equal_lavg",
      TRUE ~ NA_character_
    ),
    category_Others = case_when(
      Others > 0.5 & LQ_Others > 1 ~ "inequal_havg",
      Others > 0.5 & LQ_Others < 1 ~ "inequal_lavg",
      Others <= 0.5 & LQ_Others > 1 ~ "equal_havg",
      Others <= 0.5 & LQ_Others < 1 ~ "equal_lavg",
      TRUE ~ NA_character_
    )
  )
```


```{r}
# Relabel and prep the data
plot_df <-DFindex %>%
  filter(!is.na(category_Educational), !is.na(RUCC_2013)) %>%
  mutate(
    typology_label = recode(category_Educational,
      "inequal_havg" = "Concentrated Access",
      "inequal_lavg" = "Unequal Scarcity",
      "equal_havg"   = "Equitable Coverage",
      "equal_lavg"   = "Even Desert"
    )
  ) %>%
  group_by(Year, RUCC_2013, typology_label) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Year, RUCC_2013) %>%
  mutate(prop = n / sum(n))

# Plot
p7<-ggplot(plot_df, aes(x = Year, y = prop, fill = typology_label)) +
  geom_bar(stat = "identity", position = "fill", width = 0.8) +
  facet_wrap(~ RUCC_2013, ncol = 3) +
    scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = 2009:2021) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "(g)",
    x = "Year",
    y = "Proportion of Counties",
    fill = "Typology"
  ) +
  theme_minimal() +
  scale_fill_manual(
  values = c(
    "Equitable Coverage" = "# b2df8a",  # light green
    "Concentrated Access" = "# 1b7837", # dark green
    "Even Desert" = "# bdbdbd",         # gray
    "Unequal Scarcity" = "# 2166ac"     # dark blue
  )
)+
  theme(
    plot.title = element_text(hjust=0,face = "bold"),
    strip.text = element_text(size = 13, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1,face = "bold"),
    axis.text.y = element_text(size = 12, face = "bold"),

    axis.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )


P<- plot_grid(p1, p2, p3, p4, p5, p6,
          labels = c("(a)", "(b)", "(c)", "(d)", "(e)", "(f)"),
          ncol = 3, nrow = 2)

Pall<-plot_grid(P, p7,ncol = 1, nrow = 2)

ggsave( 'GINI Paper/Draft/Figures/County_gini_coefficient.png',dpi=300, height = 33, width = 45, units = 'cm')
```

# urbanicity topology by category (Supplementary)
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Reshape data into long format
plot_df_long <- DFindex %>%
  pivot_longer(
    cols = starts_with("category_"),
    names_to = "Service",
    values_to = "Typology"
  ) %>%
  filter(!is.na(Typology), !is.na(RUCC_2013)) %>%
  mutate(
    # Clean up service names
    Service = case_when(
      Service == "category_Educational" ~ "Educational Services",
      Service == "category_Health" ~ "Health Care",
      Service == "category_Others" ~ "Daily Living"
    ),
    # Recode typologies to descriptive labels
    Typology = recode(Typology,
      "inequal_havg" = "Concentrated Access",
      "inequal_lavg" = "Unequal Scarcity",
      "equal_havg" = "Equitable Coverage",
      "equal_lavg" = "Even Desert"
    )
  )

# Summarize counts and proportions
plot_df_summary <- plot_df_long %>%
  group_by(Year, RUCC_2013, Service, Typology) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Year, RUCC_2013, Service) %>%
  mutate(prop = n / sum(n))

p_typology_by_service <- ggplot(plot_df_summary, aes(x = Year, y = prop, fill = Typology)) +
  geom_bar(stat = "identity", position = "fill", width = 0.8) +
  facet_grid(RUCC_2013 ~ Service) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = 2009:2021) +
  scale_fill_manual(
    values = c(
      "Equitable Coverage" = "# b2df8a",      # light green
      "Concentrated Access" = "# 1b7837",     # dark green
      "Even Desert" = "# bdbdbd",             # gray
      "Unequal Scarcity" = "# 2166ac"         # dark blue
    )
  ) +
  labs(
    title = "",
    x = "Year",
    y = "Proportion of Counties",
    fill = "Typology"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0, face = "bold"),
    strip.text = element_text(size = 1, face = "bold"),
    axis.text.x = element_text(size = 18, angle = 45, hjust = 1, face = "bold"),# no
    axis.text.y = element_text(size = 18, face = "bold"), # no
    axis.title.x = element_text(size = 12, face = "bold"),  # x-axis label
    axis.title.y = element_text(size = 20, face = "bold"),  # y-axis label (edited here)

    legend.text = element_text(size = 18),
    legend.title = element_text(size = 18, face = "bold"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )


ggsave( 'GINI Paper/Draft/Figures/Supplementary_topology_by_category.png',dpi=300, height = 60, width = 35, units = 'cm')
```




```{r}
library(dplyr)
library(tidyr)

# Create a tidy version of the data by pivoting the category columns
DF_tidy <- DFindex %>%
  select(Year, State_Abbreviation, starts_with("category_")) %>%
  pivot_longer(
    cols = starts_with("category_"),
    names_to = "Sector",
    values_to = "Category"
  ) %>%
  mutate(Sector = recode(Sector,
                         category_Educational = "Educational Services",
                         category_Health = "Health Care",
                         category_Others = "Daily Living"))

# Count and calculate percentage per sector
summary_table <- DF_tidy %>%
  filter(!is.na(Category)) %>%
  group_by(Sector, Category) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Sector) %>%
  mutate(Percentage = round(Count / sum(Count) * 100, 1)) %>%
  arrange(Sector, desc(Count))

# View the table
print(summary_table)

# Optional: save to CSV for supplementary materials
write.csv(summary_table, "GINI Paper/Draft/Tables/Supplementary_Table_CaRES_Typology_countylevel.csv", row.names = FALSE)

```
